Correlation analysis
----------------------------------


```{r, echo=FALSE}
library(knitr)
if(!exists("rates_15_all")){rates_15_all <- read.csv("../Data/Tidy/rates_15_all.csv")}
if(!exists("exp_15_all")){exp_15_all <- read.csv("../Data/Tidy/exp_15_all.csv")}

# subset only 1970 onwards data for ages 20 - 50
corr.data <- merge(exp_15_all[exp_15_all$year > 1970 & 
                                exp_15_all$age >= 20 & 
                                exp_15_all$age <=50 ,],
                   rates_15_all[rates_15_all$year > 1970 &
                                  rates_15_all$age >= 20 &
                                  rates_15_all$age <=50 ,],
                   by = c("year", "age", "sex"))

# add quadrant flags
corr.data$age.split <-  cut(corr.data$age, breaks = c(20,34,50),include.lowest=TRUE, labels=c("bottom","top"))
corr.data$year.split <-   cut(corr.data$year, breaks = c(1970,1991,2012),include.lowest=TRUE, labels=c("left","right"))
corr.data$quadrants <- paste(corr.data$age.split,corr.data$year.split, sep="-" )

# helper function for calculating correlation coeficeints
FunGetCorrTable <- function(gender = "total", logged = FALSE, method = "pearson") {
  results <- matrix(NA,nrow=3, ncol=3 , 
                    dimnames = list(c("top", "bottom", "total"), c("left", "right","total")))  
  use <- corr.data[corr.data$sex ==gender,c("residual_prop", "death_rate", "quadrants", "age.split", "year.split")]
  colnames(use)[1:2] <- c("a", "b")
  if (logged == TRUE ) { use$b <- log(use$b)}
  
  results[1:2,1:2] <-matrix(by(use, use$quadrants, function(x) cor(x$a, x$b, method = method)), c(2,2), byrow = TRUE)[2:1,]
  results[3,1:2] <- by(use, use$year.split, function(x) cor(x$a, x$b, method = method))
  results[1:2,3] <- by(use, use$age.split, function(x) cor(x$a, x$b, method = method))[2:1]
  results[3,3] <- cor(use$a, use$b, method = method)
 return(results)
}

# helper function for plotting correlation coefficients
FunPlotCorrelations <- function(gender  = "total", logged = FALSE, method = "pearson") {  
  use <- corr.data[corr.data$sex ==gender , c("residual_prop", "death_rate", "year", "age")]
  colnames(use)[1:2] <- c("a", "b")
  if (logged == TRUE ) { use$b <- log(use$b)}
  par(mar=c(5.1, 4.1, 4.1, 8.1), xpd=TRUE)
  plot(0, xlim = c(1970, 2012),
  ylim = c(-1,1),
  xlab = "year",
  ylab = "correlation",
  type="n",bty='L')
  lines(unique(use$year),by(use, use$year, function(x) cor(x$a, x$b, method = method)), lwd=2, col="green")
  lines(c(1970, 2012), c(0,0), lty=2, col = "gray" )
  lines(unique(use$year),by(use[use$age >= 35,], use[use$age >= 35,"year"], function(x) cor(x$a, x$b, method = method)), lwd=2, col="blue")
 lines(unique(use$year),by(use[use$age < 35,], use[use$age < 35,"year"], function(x) cor(x$a, x$b, method = method)), lwd=2, col="red")
legend("topright", c("total", "under 35", "35 and over"), col = c("green", "blue", "red"), lty = 1, lwd =2,inset=c(-.25,0))
}

```
OK, so this is 1970 onwards, 15 European countries.
Three sets of results are reported bellow for completeness: for both sexes combined, for males only and for females only. In each case there is regular Pearson coefficients, Pearson on logged death rates and Spearman coefficients. 

Each set of results reports the four quadrants in addition to the four half-way splits and the total for the whole surface. 

I also added a year by year graph of the correlations for each set (each cobmination of log/coefficient type). These seem a lot clearer than the quadrants I'd say, showing a dramatic change after 1993, though more so for females than for males. (Not sure what's up with 2012 though.. )

## 1. Male and female combined
**Pearson** on linear data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "total", logged = FALSE, method = "pearson") 
FunPlotCorrelations(gender ="total", logged = FALSE, method = "pearson")

```

**Pearson** on **logged** data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "total", logged = TRUE, method = "pearson") 
FunPlotCorrelations(gender ="total", logged = TRUE, method = "pearson")

```

**Spearman** 
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "total", logged = FALSE, method = "spearman") 
FunPlotCorrelations(gender ="total", logged = FALSE, method = "spearman")


```

## 2. Male only 

 **Pearson** on linear data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "male", logged = FALSE, method = "pearson") 
FunPlotCorrelations(gender ="male", logged = FALSE, method = "pearson")


```

**Pearson** on **logged** data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}

FunGetCorrTable(gender = "male", logged = TRUE, method = "pearson") 
FunPlotCorrelations(gender ="male", logged = TRUE, method = "pearson")

```

**Spearman**
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "male", logged = FALSE, method = "spearman") 
FunPlotCorrelations(gender ="male", logged = FALSE, method = "spearman")


```


## 3. Female only 

**Pearson** on linear data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "female", logged = FALSE, method = "pearson") 
FunPlotCorrelations(gender ="female", logged = FALSE, method = "pearson")


```

**Pearson** on **logged** data:
```{r, echo=FALSE, fig.width = 8, fig.height = 4}

FunGetCorrTable(gender = "female", logged = TRUE, method = "pearson") 
FunPlotCorrelations(gender ="female", logged = TRUE, method = "pearson")


```

**Spearman **
```{r, echo=FALSE, fig.width = 8, fig.height = 4}
FunGetCorrTable(gender = "female", logged = FALSE, method = "spearman") 
FunPlotCorrelations(gender ="female", logged = FALSE, method = "spearman")


```


